\documentclass[11pt]{article}

\usepackage[a4paper,margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{enumitem}
\usepackage{hyperref}

\title{Consultation note: the remaining H15 quadratic pair estimate}
\author{RiemannHypothesis Lean project}
\date{July 2026}

\newcommand{\R}{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\eps}{\varepsilon}
\newcommand{\B}{\mathcal{B}}
\newcommand{\Llin}{\mathcal{L}}
\newcommand{\Lg}{\Lambda_{\gamma}}
\newcommand{\Lr}{\Lambda_{\mathrm{ratio}}}
\newcommand{\Pcent}{\mathcal{P}}
\newcommand{\fract}[1]{\left\{#1\right\}}

\begin{document}
\maketitle

\section{Purpose}

This note isolates the current H15 obstruction in the Lean formalization of the
B\'aez--Duarte/Nyman--Beurling route to RH.  The Lean development has reduced
the quadratic interaction estimate to a very specific centered
M\"obius--Bernoulli/cotangent correlation.  The aim of this consultation is not
to ask for a new Lean proof.  It is to decide the analytic status of the
remaining estimate:

\begin{quote}
Is the centered quadratic pair estimate below a known consequence of classical
Mertens/PNT-strength decay, or does it require a genuinely stronger
two-linear-forms M\"obius correlation theorem?
\end{quote}

The relevant Lean file is
\[
\texttt{RiemannHypothesis/Criteria/NymanBeurling/QuadraticInteraction.lean}.
\]
The final residual structures are
\[
\begin{array}{ll}
\texttt{QuadraticInteractionSmoothProductLogSqEstimate},&
\texttt{QuadraticInteractionCenteredBernoulliPairEstimate},\\[2mm]
\texttt{QuadraticInteractionBernoulliCorrelationEstimate},&
\texttt{quadraticInteractionEstimates\_of\_decay\_and\_bernoulliCorrelation}.
\end{array}
\]

\section{Notation}

For $1\le k\le N$, set
\[
  w_N(k)=1-\frac{k}{N+1},\qquad
  c_N(k)=-\mu(k)w_N(k),
\]
matching Lean's \texttt{cutoffMobiusCoeff}.  The relevant Vasyunin/cotangent
sum is denoted $V(h,k)$, i.e. Lean's \texttt{cotangentSumVFormula h k}.

The smooth log-gamma piece is
\[
  \Lg(N)
  =
  \sum_{h,k\le N} c_N(h)c_N(k)
  \frac{\log(2\pi)-\gamma}{2}
  \left(\frac1h+\frac1k\right).
\]
Lean proves the product identity
\[
  \Lg(N)
  =
  (\log(2\pi)-\gamma)
  \left(\sum_{h\le N}\frac{c_N(h)}h\right)
  \left(\sum_{k\le N}c_N(k)\right).
  \tag{1}
\]

The log-ratio piece is
\[
  \Lr(N)=
  \sum_{h,k\le N} c_N(h)c_N(k)
  \frac{k-h}{2hk}\log\frac hk.
\]

The BBLS Bernoulli/cotangent value is
\[
  \B(N)
  =
  -2\sum_{h,k\le N}
  \frac{c_N(h)c_N(k)}{h}
  \left(\frac{\pi}{2k}V(h,k)\right).
  \tag{2}
\]
Lean proves that this is exactly the symmetric cotangent component already used
in the quadratic decomposition:
\[
  \texttt{cutoffMobiusBernoulliCorrelationValue\_eq\_cotangentComponent}.
\]

The H14-controlled linear term is
\[
  \Llin(N)=\texttt{explicitLinearMobiusSum }N.
\]
Classical Mertens decay, formalized as
\texttt{ClassicalMertensDecay}, already gives
\[
  |\Llin(N)+1|\ll \frac1{\log(N+2)}.
\]

\section{The exact remaining estimates}

The H15 finish now asks for two estimates.  The first is a smooth-product
estimate:
\[
  |\Lg(N)|\le \frac{C_{\mathrm{smooth}}}{\log^2(N+2)}.
  \tag{S}
\]
This is \texttt{QuadraticInteractionSmoothProductLogSqEstimate}.  H14 controls
the factor $\sum c_N(h)/h$, but the currently available separate bound for
$\sum c_N(k)$ is too crude to prove (S) by multiplying absolute-value
majorants.

The second, decisive estimate is the centered Bernoulli pair estimate:
\[
  \left|
    \Lr(N)+\B(N)-1+2(\Llin(N)+1)
  \right|
  \le
  \frac{C_{\mathrm{centered}}}{\log^2(N+2)}.
  \tag{C}
\]
This is \texttt{QuadraticInteractionCenteredBernoulliPairEstimate}.  The
centering is not cosmetic: the $-1$ and doubled linear correction are the
mechanism by which the large log-ratio and cotangent/Bernoulli constituents
cancel.

Lean proves the following implications:
\[
  (S)+(C)
  \Longrightarrow
  \texttt{QuadraticInteractionBernoulliCorrelationEstimate},
\]
and, under \texttt{ClassicalMertensDecay},
\[
  \texttt{QuadraticInteractionBernoulliCorrelationEstimate}
  \Longrightarrow
  \texttt{QuadraticInteractionEstimates}.
  \tag{3}
\]
Thus the present consultation is exactly about (S) and especially (C).

\section{Lean-proved structural reductions}

Several possible sources of the main term or error have already been eliminated
or converted into exact identities.

\begin{enumerate}[label=\textbf{R\arabic*.},leftmargin=2.2em]
  \item The diagonal contribution is exactly zero:
  \[
    \texttt{quadraticInteractionDiagonal\_eq\_zero}.
  \]
  The target $+1$ does not come from the diagonal.

  \item The interaction kernel is a Gram entry minus the smooth product:
  \[
    K(h,k)
    =
    G_{h,k}
    -
    \frac{\log(2\pi)-\gamma}{2}
    \left(\frac1h+\frac1k\right).
    \tag{4}
  \]
  Consequently, for arbitrary finite weights $c$,
  \[
  \sum_{h,k\le N}c(h)c(k)K(h,k)
  =
  \int_0^\infty
  \left(
    \sum_{h\le N}c(h)\fract{\frac1{hx}}
  \right)^2dx
  -
  (\log(2\pi)-\gamma)
  \left(\sum_{h\le N}\frac{c(h)}h\right)
  \left(\sum_{k\le N}c(k)\right).
  \tag{5}
  \]
  This is
  \texttt{quadratic\_form\_eq\_normSq\_sub\_linear}.

  \item Non-squarefree gcd slices vanish identically:
  \[
    \texttt{quadraticInteractionGcdSlice\_eq\_zero\_of\_not\_squarefree}.
  \]
  This validates gcd stratification but does not by itself control the
  squarefree strata.

  \item The BBLS Bernoulli/Fourier representation is proved as a partial-sum
  limit:
  \[
    \texttt{cotangent\_component\_eq\_mode\_partial\_limit}.
  \]
  Lean deliberately does not claim absolute/unordered summability of the mode
  series without an explicit summability hypothesis.
\end{enumerate}

\section{Numerical negatives that shape the question}

The diagnostics below were re-run in this worktree using the bundled Python
runtime.  They are not proofs, but they prevent us from freezing the wrong
analytic target.

\subsection{Per-mode bounds do not look viable}

Writing the cotangent/Bernoulli component as
\[
  C(N)=\sum_{j\ge1}W_j(N),
\]
the first mode is stable near $1$, but the remaining absolute mass does not
decay in a way compatible with a simple per-mode estimate.

\begin{center}
\begin{tabular}{c|c|c|c}
$N$ & $W_1(N)$ & $\sum_{1\le j\le8}|W_j(N)|$
& $\sum_{9\le j\le96}|W_j(N)|$\\
\hline
300  & $1.01533$ & $1.48639$ & $3.11909$\\
1000 & $1.03192$ & $1.55252$ & $3.28921$
\end{tabular}
\end{center}

The exact small-$N$ check of $S_2=C-W_1$ also does not show decay:
\[
\begin{array}{c|c|c|c}
N & C(N) & W_1(N)-1 & S_2(N)\\
\hline
100 & 0.74471 & -0.03951 & -0.21578\\
200 & 1.14859 &  0.00505 &  0.14354\\
400 & 0.80217 &  0.02285 & -0.22068
\end{array}
\]
So the estimate should not be split as ``$W_1-1$ plus a decaying mode tail''.
The cancellation appears to couple the non-$j=1$ modes with the log-ratio,
log-gamma, and linear correction terms.

\subsection{The centered package is small}

The Lean comments record the decisive centered numerics:
\[
  \left(
    |\Lg(N)|+
    \left|\Lr(N)+\B(N)-1+2(\Llin(N)+1)\right|
  \right)\log^2(N+2)
\]
takes the values
\[
  4.01,\ 3.11,\ 2.38,\ 1.56
  \qquad (N=20,50,100,300).
\]
For the centered pair alone, Lean records:
\[
  \left|\Lr(N)+\B(N)-1+2(\Llin(N)+1)\right|\log^2(N+2)
  \approx
  0.975,\ 0.925,\ 1.010
\]
at $N=500,1000,2000$.

\section{The consultation question}

Please assess the following statement.

\begin{quote}
Assume classical Mertens decay, for instance
\[
  M(x)=\sum_{n\le x}\mu(n)
  \ll x\exp(-a\sqrt{\log x})
\]
with some $a>0$, plus the standard consequences already formalized in H14.
Do the estimates (S) and (C) follow?
\end{quote}

More specifically:

\begin{enumerate}[label=\textbf{Q\arabic*.},leftmargin=2.2em]
  \item Can (S) be proved by a classical bilinear/Abel argument controlling the
  product in (1), or does it require a genuinely two-factor cancellation not
  implied by separate bounds?

  \item Is (C) a known consequence of BBLS/Estermann autocorrelation theory,
  Vasyunin cotangent-sum identities, or standard de la Vall\'ee
  Poussin/Perron technology?

  \item If (C) is not a consequence of one-variable Mertens decay, what is the
  minimal known arithmetic hypothesis that implies it?  In particular, does it
  require a uniform two-linear-forms correlation estimate of schematic type
  \[
    \sum_{m\le M}
      \mu(a m+r)\mu(q m+s)\,W(m)
    \ll
    \frac{\|W\|_1}{\log^A N},
  \]
  summed over Farey-cell parameters?

  \item Is there an alternative centering identity that makes the $+1$ main
  term and the correction $2(\Llin(N)+1)$ automatic, avoiding a two-linear
  forms hypothesis?

  \item If the displayed estimate is false or too strong, what asymptotic
  should replace it so that the Lean reduction remains mathematically honest?
\end{enumerate}

\section{What would be enough for Lean}

For the formal project, a satisfactory external answer can be one of:

\begin{enumerate}[label=\textbf{A\arabic*.},leftmargin=2.2em]
  \item a proof of (S) and (C) from \texttt{ClassicalMertensDecay};
  \item a proof from a precisely stated classical theorem stronger than
  \texttt{ClassicalMertensDecay};
  \item a correction showing that (C) needs a new named hypothesis, together
  with the cleanest mathematically standard statement of that hypothesis;
  \item a counter-asymptotic showing that the current centered estimate should
  be weakened or re-centered.
\end{enumerate}

The Lean side is already prepared for option A1 or A2 through
\[
\texttt{quadraticInteractionEstimates\_of\_decay\_and\_bernoulliCorrelation}.
\]
If the answer is A3, the project should replace
\texttt{QuadraticInteractionCenteredBernoulliPairEstimate} by the sharper named
correlation package suggested by the consultation.

\section{Current Lean ledger after this consultation}

The verified baseline \texttt{verified-h15-consultation-note} contains this
consultation note but not the final A6 proof.  Consequently no theorem named
\texttt{riemannHypothesis\_of\_pair\_and\_NBforward} is stated from this note
alone.  The intended final assembly is still gated by the centered pair
estimate above: once that estimate is supplied, the existing Lean bridges are
designed to route it through
\texttt{quadraticInteractionEstimates\_of\_decay\_and\_bernoulliCorrelation}
and the Nyman--Beurling forward bridge, or the current NB/RH axiom.

The generator hygiene performed after the consultation is intentionally
orthogonal to the mathematics: future certificate emitters now insert
\texttt{set\_option linter.style.longLine false} headers, but no generated
certificate file was regenerated and no public definition value was changed.

\end{document}
