\documentclass[11pt,a4paper]{article}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage[margin=2.6cm]{geometry}
\usepackage{hyperref}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}
\newcommand{\fract}[1]{\{#1\}}
\title{A Machine-Verified Reduction of the Nyman--Beurling--B\'aez-Duarte
Criterion to Two Classical Statements\\[1ex]
\large An LLM-guided formalization pipeline in Lean~4 --- a working route, not a proof}
\author{Xavier Fresquet \and G\'erard Biau\\[0.5ex]
\small Sorbonne Universit\'e}
\date{July 2026}
\begin{document}
\maketitle

\begin{abstract}
We report on an experiment in AI-assisted mathematics. Large language models
were used to \emph{navigate} the classical literature around the
Nyman--Beurling--B\'aez-Duarte (NB--BD) criterion for the Riemann Hypothesis,
proposing decompositions, catching false statements, and drafting proofs;
the Lean~4 proof assistant (with Mathlib) was used as the sole arbiter of
correctness. \textbf{We do not claim a proof of RH, nor of any new theorem of
independent strength.} The contribution is methodological: starting from a
trust-nothing state (1{,}364 unaudited axioms), the pipeline has produced a
fully machine-checked reduction in which the B\'aez-Duarte criterion follows
from exactly \emph{two} precisely-stated analytic assertions --- one being the
classical 1899 de la Vall\'ee Poussin Mertens bound, the other a single
numerically-supported correlation estimate --- with every reduction step
verified by the Lean kernel (\texttt{\#print axioms} audits throughout).
We describe the route, the key identities that were formalized along the way
(several of which required correcting false statements caught numerically),
and the two open statements, in the hope that a professional analytic number
theorist may recognize the remaining distance --- or close it.
\end{abstract}

\section{Framing: what this is and is not}
The role of the LLMs here is \emph{navigation}: surveying candidate routes
(digamma identities, Estermann zeta contours, Farey-cell correlations,
iterated cotangent reciprocity), discarding dead ends cheaply, and proposing
the next well-scoped lemma. The role of Lean is \emph{rigour}: no step enters
the trusted corpus unless the kernel accepts it, and every claimed theorem is
audited down to its axioms. Numerical verification precedes every
formalization step; this discipline caught three false statements that would
otherwise have poisoned the development (\S\ref{sec:false}).
Nothing below should be read as a claim on RH. The honest summary is:
\emph{a certified reduction with two explicit, named debts}.

\section{The criterion and the Vasyunin Gram matrix}
Let $\rho_k(x)=\fract{1/(kx)}$ on $(0,\infty)$ and $\chi=\mathbf 1_{(0,1]}$.
B\'aez-Duarte (2003): RH holds iff $\chi$ lies in the $L^2(0,\infty)$-closure
of $\mathrm{span}\{\rho_k : k\ge 1\}$. The finite-$N$ analysis rests on the
Gram integrals
\[
G_{h,k}=\int_0^\infty \fract{\tfrac1{hx}}\fract{\tfrac1{kx}}\,dx ,
\]
for which Vasyunin's explicit formula holds (now a Lean theorem, see below):
\[
G_{h,k}
=\frac{\log 2\pi-\gamma}{2}\Bigl(\frac1h+\frac1k\Bigr)
+\frac{k-h}{2hk}\log\frac hk
-\frac{\pi}{2hk}\bigl(V(h,k)+V(k,h)\bigr),
\qquad
V(h,k)=\sum_{a=1}^{k-1}\fract{\tfrac{ah}{k}}\cot\frac{\pi a}{k}.
\]

\section{What is machine-proved (selected milestones)}
\subsection{The Vasyunin bridge (closed)}
Following B\'aez-Duarte--Balazard--Landreau--Saias (Ramanujan J.\ 9 (2005);
arXiv:math/0306251), the full period-reduction chain was formalized:
their Propositions 12, 15, 16 (a Frullani-type identity, proved for all real
$\theta,x>0$), 21 (rational case), 22 (the Stieltjes identity), 48, 87, 88, 89.
Two representative statements, both Lean theorems with clean axiom audits:
\[
\varphi_1\!\Bigl(\frac pq\Bigr):=\sum_{k\ge1}\frac{B_1(kp/q)}{k}
=\frac{\pi}{2q}V(p,q)
\quad\text{(all $p,q\ge1$, \emph{no} coprimality)},
\qquad
A(1)=\int_0^\infty\frac{\fract t^2}{t^2}\,dt=\log 2\pi-\gamma,
\]
where $B_1$ is the Dedekind sawtooth. The $x\to\infty$ limit of Proposition~22
gives the autocorrelation $A(k/h)$ in closed cotangent form, whence the
period reduction and Vasyunin's formula for $G_{h,k}$ ($h,k\ge1$)
unconditionally. Notably, the digamma reflection input was rebuilt from a
real-variable Mittag-Leffler cotangent expansion, avoiding any digamma API.

\subsection{The linear (M\"obius--Dirichlet) debt: collapsed to one classical bound}
Define the single hypothesis
\[
\textsf{ClassicalMertensDecay}:\qquad
\exists\,C,a>0:\ |M(N)|=\Bigl|\sum_{k\le N}\mu(k)\Bigr|\le C\,N\,e^{-a\sqrt{\log N}}
\quad(N\ge2),
\]
i.e.\ exactly the output of de la Vall\'ee Poussin's 1899 zero-free region.
The following are now Lean theorems \emph{conditional only on this}:
the quantitative bounds $|M(N)|\le C_M N/\log^3(N{+}2)$ and
$|L(N)|\le C_L N/\log^2(N{+}2)$ for $L(N)=\sum_{k\le N}\mu(k)\log k$;
a quantitative Axer normalization
\[
\Bigl|\sum_{k\le N}\frac{\mu(k)}{k}\Bigr|\le C' e^{-c'\sqrt{\log N}}
\]
(via the hyperbola identity $\sum_{k\le N}\mu(k)\lfloor N/k\rfloor=1$ and a
blockwise Abel argument on the monotone pieces of $k\mapsto\fract{N/k}$); and
the limit $\sum_{k\le N}\mu(k)\log k/k\to-1$, obtained \emph{without} PNT
normalization by an Abelian boundary theorem for Dirichlet series combined
with $\sum\mu(n)\log n\,n^{-s}=\zeta'(s)/\zeta(s)^2\to-1$ as $s\to1^+$,
anchored at the pole via the analytic function $(s-1)\zeta(s)$. In Lean:
\[
\texttt{ClassicalMertensAPI.ofDecayOnly}:\ \textsf{ClassicalMertensDecay}
\longrightarrow \texttt{ClassicalMertensAPI}.
\]

\subsection{The quadratic (interaction) debt: reduced to one centered estimate}
The interaction kernel of the quadratic remainder turned out to be, literally,
\[
K(h,k)=G_{h,k}-\frac{\log2\pi-\gamma}{2}\Bigl(\frac1h+\frac1k\Bigr),
\]
so that for any weights $c$,
\[
\sum_{h,k\le N}c(h)c(k)K(h,k)
=\int_0^\infty\Bigl(\sum_{h\le N}c(h)\fract{\tfrac1{hx}}\Bigr)^{\!2}dx
-(\log2\pi-\gamma)\Bigl(\sum_{h\le N}\frac{c(h)}{h}\Bigr)\Bigl(\sum_{k\le N}c(k)\Bigr),
\]
a manifestly positive-semidefinite norm term minus a product of linear sums
(both identities are Lean theorems). Numerical diagnostics had shown that
per-gcd-stratum bounds are individually false; the norm form explains the
observed global cancellation structurally. An exact BBLS-style triple-series
rearrangement of the M\"obius-weighted cotangent sum is also proved, together
with the \emph{negative} finding that its pointwise inner sums grow
($\approx2.9\to14.6$ for $N=10^2\to3\cdot10^3$): only the \emph{centered}
package can be bounded. The remaining debt is the single field
\[
\bigl|\Lambda_\gamma(N)\bigr|
+\bigl|\Lambda_{\mathrm{ratio}}(N)+\mathcal B(N)-1+2(\mathcal L(N)+1)\bigr|
\;\le\;\frac{C}{\log^2(N{+}2)},
\tag{$\ast$}
\]
where $\mathcal B$ is a cutoff M\"obius--Bernoulli correlation value,
$\mathcal L$ the explicit linear M\"obius sum, and
$\Lambda_\gamma,\Lambda_{\mathrm{ratio}}$ the log-gamma/log-ratio components.
Numerically $(\ast)\cdot\log^2(N{+}2)$ decreases $4.0\to1.6$ over
$N=20\to300$. Lean-proved bridges give: \textsf{ClassicalMertensDecay}
$+\;(\ast)\;\Longrightarrow$ the full quadratic estimates.

\section{Falsehoods caught by the pipeline}\label{sec:false}
Three statements in the development were \emph{false as recorded} and caught
by mandatory numerics or by the kernel: an interval-tsum identity for
$G_{h,k}$ missing its tail; an axiomatized rational upper bound for $\pi$
that was strictly below $\pi$; and the hypothesis-free form of Vasyunin's
formula, false at $h=0$ (the Gram integrand vanishes identically while the
formula does not). Each was corrected, re-verified, and its $\sim$2{,}000
call sites mechanically patched. We consider this catch-rate the strongest
argument for the methodology.

\section{Current state and the two open statements}
The entire pipeline --- Vasyunin bridge, linear estimates, quadratic
estimates, finite PSD certificates up to $N=50$ --- is machine-checked, with
exactly one decorative \texttt{sorry} (the general-real-$\theta$ form of BBLS
Proposition~21, used by nothing) and a short, audited axiom list. The
B\'aez-Duarte criterion is a Lean theorem conditional on precisely:
\begin{enumerate}
\item \textsf{ClassicalMertensDecay} --- classical (dVP 1899); its
formalization is staged (the $3$-$4$-$1$ positivity ingredients, von Mangoldt
cosine expansion, and local zeta wrappers are already proved) but the
zero-free region instance and effective Perron argument remain a substantial
classical formalization project;
\item the centered correlation estimate $(\ast)$ --- to our knowledge not a
named theorem in the literature; numerically robust; we would be grateful for
expert opinion on whether it follows from known results (Landau/Axer-style
normalizations, Estermann zeta growth, or correlation estimates of
Matom\"aki--Radziwi\l\l--Tao type) or constitutes genuinely open territory.
\end{enumerate}
Beyond these sits the NB--BD equivalence itself (currently an axiom in the
development; only the forward direction criterion~$\Rightarrow$~RH would be
needed, via a Mellin vanishing argument).

\section*{Invitation}
All Lean sources, verification scripts, and numerical diagnostics are
available for inspection; every theorem named above can be audited with
\texttt{\#print axioms} in minutes. We would value a mathematician's
judgement on statement $(\ast)$ above all.

\begin{thebibliography}{9}\small
\bibitem{BD} L.~B\'aez-Duarte, \emph{A strengthening of the Nyman--Beurling
criterion for the Riemann hypothesis}, Rend.\ Lincei Mat.\ Appl.\ 14 (2003).
\bibitem{BBLS} L.~B\'aez-Duarte, M.~Balazard, B.~Landreau, E.~Saias,
\emph{\'Etude de l'autocorr\'elation multiplicative de la fonction «partie
fractionnaire»}, Ramanujan J.\ 9 (2005); arXiv:math/0306251.
\bibitem{Vas} V.~I. Vasyunin, \emph{On a biorthogonal system associated with
the Riemann hypothesis}, St.~Petersburg Math.\ J.\ 7 (1996).
\bibitem{dVP} C.-J. de la Vall\'ee Poussin, M\'em.\ Couronn\'es Acad.\ Roy.\
Belgique 59 (1899).
\bibitem{Lean} The mathlib Community, \emph{The Lean mathematical library},
CPP 2020; L.~de Moura, S.~Ullrich, CADE-28 (2021).
\end{thebibliography}
\end{document}
